---
title: "session 4 notes"
author: "Derek Sollberger"
format: html
---

```{r}
library("tidyverse")

demo_df <- readr::read_csv("https://raw.githubusercontent.com/dsollberger/sml201slides/main/posts/04_categories/sml201survey.csv")
```

# Grouping

How many languages do Princeton students speak at each residential college?

* here: population is all survey respondents

```{r}
# population mean
mu <- mean(demo_df$languages, na.rm = TRUE)
mu
```

## Filtering

```{r}
df2 <- select(demo_df, residentialCollege, languages)

df3 <- filter(df2, residentialCollege == "Mathey")

# sample means
xbar <- mean(df3$languages, na.rm = TRUE)
xbar

df4 <- filter(df2, residentialCollege == "Forbes")
xbar <- mean(df4$languages, na.rm = TRUE)
xbar
```

* key observation: the sample mean can change based on the sampling

```{r}
df_grouped <- group_by(df2, residentialCollege)
summarize(df_grouped,
          xbar = mean(languages, na.rm = TRUE))
```

```{r}
all_means <- summarize(df_grouped,
          xbar = mean(languages, na.rm = TRUE))

# variables go inside aesthetics
ggplot(all_means, aes(x = residentialCollege, y = xbar)) +
  
  # when we want to plot the numbers directly, use "identity"
  geom_bar(color = "red", fill = "green",stat = "identity")
```

# Pipes

The pipe (`|>`) inputs what comes before into the first argument of the next function.

```{r}
demo_df |>
  select(residentialCollege, languages) |>
  group_by(residentialCollege) |>
  summarise(xbar = mean(languages, na.rm = TRUE)) |>
  ungroup() |>
  ggplot(aes(x = residentialCollege, y = xbar)) +
  geom_bar(colour = "red", fill = "green",stat = "identity")
```

What are the most frequent majors among SML 201 students?

```{r}
df2 <- demo_df |>
  select(major) |>
  group_by(major) |>
  mutate(numMajor = n()) |>  #counts rows per group
  ungroup() |>
  distinct() |>
  arrange(desc(numMajor))
```

```{r}
df3 <- demo_df |>
  select(major) |>
  count(major) |>
  arrange(desc(n))
```

```{r}
demo_df |>
  ggplot(aes(x = fct_rev(fct_infreq(major)),
             fill = major)) +
  coord_flip() +
  geom_bar(stat = "count") +
  labs(title = "Majors of SML 201 Students",
       subtitle = "Fall 2024",
       y = "number of students",
       x = "") +
  theme(legend.position = "none")
```


## More Examples

What time do you go to sleep?

```{r}
table(demo_df$sleepTime)
```

```{r}
summary(demo_df$flossing)
```

```{r}
summary(demo_df$attractiveness)
```

```{r}
summary(demo_df$intelligence)
```

```{r}
summary(demo_df$showers)
```
```{r}
table(demo_df$drugUse)
```

```{r}
table(demo_df$pancakesWaffles)
```

```{r}
summary(demo_df$SAT)

demo_df$SAT[demo_df$SAT > 1600] <- NA
demo_df$SAT[demo_df$SAT < 400] <- NA

summary(demo_df$SAT)
```










