---
title: "22: Unsupervised Learning"
author: "Derek Sollberger"
date: "2026-04-21"
format:
  html:
    toc: true
    theme: cerulean
---

# SML 201

## Start

::: {.callout-note collapse="true"}
## Libraries and Loading the Data

```{r}
#| message: false
#| warning: false

library("corrplot")
library("DiagrammeR")#easy coding of flowcharts
library("GGally")    #extensions for ggplot
library("ggsignif")  #show significance levels on boxplots
library("gt")
library("patchwork") #arrange plots
library("sf")
library("tidyclust")
library("tidymodels")
library("tidyverse") #general framework for data wrangling

# school colors
princeton_orange <- "#E77500"
princeton_black  <- "#121212"

# data set: College Scorecard
# college_raw <- readr::read_csv('Most-Recent-Cohorts-Institution.csv')
# college_df <- college_raw |>
#   select(INSTNM, UGDS,
#          STABBR, LATITUDE, LONGITUDE,
#          ADM_RATE,
#          # SATMT75, SATWR75,
#          C150_4, RET_FT4,
#          MD_EARN_WNE_P10, ENDOWBEGIN) |>
#   drop_na() |>
#   filter(LONGITUDE > -125 & LONGITUDE < -67) |>
#   filter(LATITUDE > 25 & LATITUDE < 49)
# readr::write_csv(college_df, "college_data.csv")
college_df <- readr::read_csv("college_data.csv")
# college_df <- readr::read_csv("https://raw.githubusercontent.com/dsollberger/sml201slides/refs/heads/main/data/college_data.csv")
states_shp <- readr::read_rds("us_states_shp.rds")
# states_shp <- readr::read_rds("https://raw.githubusercontent.com/dsollberger/sml201slides/refs/heads/main/data/us_states_shp.rds")

set.seed(201)
control_data <- data.frame(
  age = sample(18:65, size = 100, replace = TRUE),
  height = round(rnorm(100, 65, 8)),
  sex = sample(c("female", "male"), size = 100, replace = TRUE),
  group = "control",
  weight_before = round(rnorm(100, 150, 15))
) |>
  #coded like Gaussian white noise
  mutate(weight_after = round(weight_before - rnorm(100, 0, 6)))

treatment_data <- data.frame(
  age = sample(18:65, size = 100, replace = TRUE),
  height = round(rnorm(100, 65, 8)),
  sex = sample(c("female", "male"), size = 100, replace = TRUE),
  group = "treatment",
  weight_before = round(rnorm(100, 150, 15))
) |>
  #coded like Gaussian white noise
  mutate(weight_after = round(weight_before - rnorm(100, 15, 6)))

GLP_raw <- rbind(control_data, treatment_data) |>
  sample_frac() |>
  mutate(patient_id = 1:200, .before = "age")

# helper function
vnorm <- function(x, mu = 0, sigma = 1, section = "lower"){
  
  # bell curve
  x_vals <- seq(mu - 4*sigma, mu + 4*sigma, length.out = 201)
  y_vals <- dnorm(x_vals, mu, sigma)
  df_for_graph <- data.frame(x_vals, y_vals)

  # outline shaded regions
  if(length(x) == 1){
    shade_left <- rbind(c(x[1],0), df_for_graph |>
                        filter(x_vals < x[1]))
    shade_right <- rbind(c(x[1],0), df_for_graph |>
                        filter(x_vals > x[1]))
  }
  if(length(x) == 2){
    shade_between <- rbind(c(x[1],0),
                       df_for_graph |>
                         filter(x_vals > x[1] &
                                  x_vals < x[2]),
                       c(x[2],0))
    shade_tails <- rbind(df_for_graph |>
                        filter(x_vals < x[1]),
                     c(x[1],0),
                     c(x[2],0),
                     df_for_graph |>
                        filter(x_vals > x[2]))
  }
  if(section == "lower"){
    bell_curve <- df_for_graph |>
      ggplot(aes(x_vals, y_vals)) +
      geom_polygon(aes(x = x_vals, y = y_vals),
                   data = shade_left,
                   fill = "#E77500",) +
      geom_line(color = "gray50", linewidth = 2)
    prob_val <- round(pnorm(x,mu,sigma), 4)
  }
  if(section == "upper"){
    bell_curve <- df_for_graph |>
      ggplot(aes(x_vals, y_vals)) +
      geom_polygon(aes(x = x_vals, y = y_vals),
                   data = shade_right,
                   fill = "#E77500",) +
      geom_line(color = "gray50", linewidth = 2)
    prob_val <- 1 - round(pnorm(x,mu,sigma), 4)
  }
  if(section == "between"){
    bell_curve <- df_for_graph |>
      ggplot(aes(x_vals, y_vals)) +
      geom_polygon(aes(x = x_vals, y = y_vals),
                   data = shade_between,
                   fill = "#E77500",) +
      geom_line(color = "gray50", linewidth = 2)
    prob_val <- round(diff(pnorm(x,mu,sigma)), 4)
  }
  if(section == "tails"){
    bell_curve <- df_for_graph |>
      ggplot(aes(x_vals, y_vals)) +
      geom_polygon(aes(x = x_vals, y = y_vals),
                   data = shade_tails,
                   fill = "#E77500",) +
      geom_line(color = "gray50", linewidth = 2)
    prob_val <- round(1 - diff(pnorm(x,mu,sigma)), 4)
  }
  
  # plot bell curve
  bell_curve + 
    labs(subtitle = paste0("Probability: ", prob_val),
         caption = "SML 201", y = "") +
    theme_minimal()
}

```

:::

:::: {.columns}

::: {.column width="50%"}
* **Goal**: Explore topics in unsupervised learning


:::

::: {.column width="10%"}

:::

::: {.column width="40%"}
* **Objectives**: 

- clustering
- workflows

:::

::::

## Data: College Scorecard

Today's data comes from [College Scorecard](https://collegescorecard.ed.gov/data).

* `INSTNM`: institution name
* `UGDS`: undergraduate enrollment (number of students)
* `LATITUDE`
* `LONGITUDE`
* `ADM_RATE`: admission rate
* `C150_4`: completion rate (at 150% of expected time for 4-year degrees)
* `RET_FT4`: retention rate (full-time students, 4-year institutions)
* `MD_EARN_WNE_P10`: median earnings, 10 years after graduation (USD)
* `ENDOWBEGIN`: endowment (beginning of school year)

## Correlation Plot

```{r}
college_df |>
  select_if(is.numeric) |>
  cor(use = "pairwise.complete.obs") |>
  corrplot(method = "ellipse",
           order = "FPC",
           type = "upper")
```




# k-means Clustering

## Scene

```{r}
states_shp |>
  ggplot() +
  geom_sf(color = "gray80", fill = "white") +
  geom_point(aes(x = LONGITUDE, y = LATITUDE, 
                 color = STABBR),
             data = college_df) +
  labs(title = "American Universities",
       subtitle = "A selection of insitutions in the continental USA",
       caption = "Source: College Scorecard",
       x = "", y = "") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

## A first foray

```{r}
set.seed(20241121)
k4_fit <- k_means(num_clusters = 4) |>
  set_engine("stats") |>
  fit(~ LATITUDE + LONGITUDE, data = college_df)
```

## Centroids

```{r}
states_shp |>
  ggplot() +
  geom_sf(color = "gray80", fill = "white") +
  geom_point(aes(x = LONGITUDE, y = LATITUDE),
             color = "gray60",
             data = college_df) +
  geom_point(aes(x = LONGITUDE, y = LATITUDE, 
                 color = .cluster),
             data = extract_centroids(k4_fit),
             size = 7) +
  labs(title = "Centroids",
       subtitle = "The clusters will be picked by distance to the nearest centroid",
       caption = "Source: College Scorecard",
       x = "", y = "") +
  scale_color_brewer(palette = "Dark2") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

## Clusters

```{r}
cluster_df <- college_df |>
  mutate(cluster_num = extract_cluster_assignment(k4_fit) |>
           pull())

states_shp |>
  ggplot() +
  geom_sf(color = "gray80", fill = "white") +
  geom_point(aes(x = LONGITUDE, y = LATITUDE, 
                 color = cluster_num),
             data = cluster_df) +
  labs(title = "Clusters",
       subtitle = "k = 4",
       caption = "Source: College Scorecard",
       x = "", y = "") +
  scale_color_brewer(palette = "Dark2") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

::: {.callout-note}
## Considerations

1. How do we place the centroids? Randomly?
2. How do we pick the number of clusters?
:::




# Error Ratios

## Try several arrangements

```{r}
N <- 10 #max possible number of clusters
k_vals <- 1:N
err_ratios <- rep(NA, N)

for(i in 1:N){
  kmeans_fit <- k_means(num_clusters = k_vals[i]) |>
    set_engine("stats") |>
    fit(~ LATITUDE + LONGITUDE, data = college_df)
  err_ratios[i] <- kmeans_fit$fit$tot.withinss / kmeans_fit$fit$totss
}
```

## Scree Plot

```{r}
df_for_graph <- data.frame(num_clusters = 1:N,
                           err_ratios)

df_for_graph |>
  ggplot() +
  geom_point(aes(x = factor(num_clusters), y = err_ratios),
             size = 3) +
  geom_line(aes(x = num_clusters, y = err_ratios)) +
  labs(title = "Scree Plot",
       subtitle = "Aiming to choose the number of clusters",
       caption = "SML 201",
       x = "number of clusters",
       y = "within/total SSE ratio") +
  theme_minimal(base_size = 14)
```

::: {.callout-warning}
## DCP1
:::


# Workflow

::::: {.panel-tabset}

## Flowchart

```{r}
#| echo: false
#| eval: true
DiagrammeR::mermaid("
graph LR

data_frame --> cross_validation
data_frame --> recipe
recipe --> workflow
spec --> workflow
grid_search --> tuning
workflow --> tuning
cross_validation --> tuning

tuning --> eval_metrics
")
```

## Mermaid code

```{r}
#| echo: true
#| eval: false
DiagrammeR::mermaid("
graph LR

data_frame --> cross_validation
data_frame --> recipe
recipe --> workflow
spec --> workflow
grid_search --> tuning
workflow --> tuning
cross_validation --> tuning

tuning --> eval_metrics
")
```

:::::

## Tidymodels

```{r}
# cross-validation folds
college_cv <- vfold_cv(college_df, v = 10)

# specification
college_spec <- k_means(num_clusters = tune())

# recipe
college_recipe <- recipe(~., data = college_df |>
                         select_if(is.numeric)) |>
  step_normalize()

# workflow
college_workflow <- workflow(college_recipe, college_spec)

# parameter grid
kvals_grid <- grid_regular(num_clusters(), levels = 10)

# tuning
tune_results <- tune_cluster(
  college_workflow,
  resamples = college_cv,
  grid = kvals_grid,
  metrics = cluster_metric_set(sse_within_total, sse_total, sse_ratio)
)
```

## Scree Plot

```{r}
tune_results |>
  collect_metrics() |>
  filter(.metric == "sse_ratio") |>
  ggplot(aes(x = num_clusters, y = mean)) +
  geom_point(aes(x = factor(num_clusters), y = mean),
             size = 3) +
  geom_line() +
  labs(title = "Scree Plot",
       subtitle = "Aiming to choose the number of clusters",
       caption = "SML 201",
       x = "number of clusters",
       y = "within/total SSE ratio") +
  theme_minimal(base_size = 14)
```

# Pairs Plot

```{r}
#| message: false
college_df |>
  select_if(is.numeric) |>
  GGally::ggpairs(title = "American Universities")
```

```{r}
set.seed(20241121)
k4_fit <- k_means(num_clusters = 4) |>
  set_engine("stats") |>
  fit(~., data = college_df |>
        select_if(is.numeric))
college_k4 <- college_df |>
  mutate(cluster_num = extract_cluster_assignment(k4_fit) |>
           pull())
```

```{r}
#| message: false
college_k4 |>
  GGally::ggpairs(columns = c(2, 6:9),
                  aes(color = cluster_num)) +
  scale_color_brewer(palette = "Dark2")
```




# Classification

Out of interest, we now seek smaller cluster sizes (i.e. fewer colleges in each cluster), so let us explore an arrangement with $k = 10$ clusters.

```{r}
set.seed(20241121)
k10_fit <- k_means(num_clusters = 10) |>
  set_engine("stats") |>
  fit(~., data = college_df |>
        select_if(is.numeric) |>
        mutate(across(where(is.numeric), scale)))
college_k10 <- college_df |>
  mutate(cluster_num = extract_cluster_assignment(k10_fit) |>
           pull())
```

```{r}
p_clust_id <- college_k10 |> 
  filter(INSTNM == "Princeton University") |>
  pull(cluster_num)
```

```{r}
college_k10 |>
  filter(cluster_num == p_clust_id)
```
::::: {.panel-tabset}

## table

```{r}
#| echo: false
#| eval: true
college_k10 |>
  filter(cluster_num == p_clust_id) |>
  select(INSTNM, STABBR, UGDS,
         ADM_RATE, C150_4, RET_FT4,
         MD_EARN_WNE_P10) |>
  rename(university = INSTNM,
         state = STABBR,
         enrollment = UGDS,
         income = MD_EARN_WNE_P10) |>
  mutate(admit_rate = 100*ADM_RATE,
         grad_rate = 100*C150_4,
         retention = 100*RET_FT4,) |>
  select(university, state, enrollment,
         admit_rate, grad_rate, retention,
         income) |>
  arrange(university) |>
  gt() |>
  cols_align(align = "center") |>
  tab_footnote(footnote = "Source: College Scorecard") |>
  tab_header(
    title = "The Princeton Cluster",
    subtitle = "of similar universities"
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = list(cell_fill(color = "#E77500"),
                 cell_text(color = "#121212",
                           weight = "bold")),
    locations = cells_body(rows = university == "Princeton University")
  )
```

## gt

```{r}
#| echo: true
#| eval: false
college_k10 |>
  filter(cluster_num == p_clust_id) |>
  select(INSTNM, STABBR, UGDS,
         ADM_RATE, C150_4, RET_FT4,
         MD_EARN_WNE_P10) |>
  rename(university = INSTNM,
         state = STABBR,
         enrollment = UGDS,
         income = MD_EARN_WNE_P10) |>
  mutate(admit_rate = 100*ADM_RATE,
         grad_rate = 100*C150_4,
         retention = 100*RET_FT4,) |>
  select(university, state, enrollment,
         admit_rate, grad_rate, retention,
         income) |>
  arrange(university) |>
  gt() |>
  cols_align(align = "center") |>
  tab_footnote(footnote = "Source: College Scorecard") |>
  tab_header(
    title = "The Princeton Cluster",
    subtitle = "of similar universities"
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = list(cell_fill(color = "#E77500"),
                 cell_text(color = "#121212",
                           weight = "bold")),
    locations = cells_body(rows = university == "Princeton University")
  )
```


:::::

::: {.callout-warning}
## DCP2
:::


# Review Session

## Data

> "GLP [Glucagon-like peptide-1] agonists are medications that help lower blood sugar levels and promote weight loss." --- Cleveland Clinic

We are going to take hypothetical data and pretend that we are analyzing results from clinical trials in hopes of obtaining approval from the Food and Drug Administration.


### Data Preparation

* add a column for

$$\text{weight loss} = \text{weight before} - \text{weight after}$$

* remove the `patient_id` column

```{r}
GLP_df <- GLP_raw |>
  mutate(weight_loss = weight_before - weight_after) |>
  select(-patient_id)
```

## Probability

Assuming a normal distribution, what is the probability that a randomly selected person in the treatment group lost at least 10 pounds?

```{r}
GLP_df |>
  filter(group == "treatment") |>
  summarize(xbar = mean(weight_loss),
            s = sd(weight_loss))

pnorm(10, 16.1, 5.8232, lower.tail = FALSE)
```

```{r}
vnorm(10, 16.1, 5.8232, section = "upper") +
  labs(title = "GLP Case Study",
       subtitle = "probability of losing at least 10 pounds",
       caption = "SML 201\n(hypothetical data)",
       x = "weight loss (pounds)")
```


## Confidence Intervals

Use `infer::visualize` to construct confidence intervals for `weight_loss` for both the control and treatment groups.

```{r}
bootstrap_distribution <- GLP_df |>
  filter(group == "control") |>
  specify(response = weight_loss) |>
  generate(reps = 1337, type = "bootstrap") |>
  calculate(stat = "mean")

bootstrap_distribution |>
  visualize() +
  shade_ci(endpoints = bootstrap_distribution |>
             get_ci(level = 0.95, type = "percentile"),
           color = princeton_black,
           fill = princeton_orange) +
  labs(title = "GLP Clinical Trials",
       subtitle = "control group",
       caption = "(hypothetical data)",
       x = "weight loss", y = "number of patients") +
  theme_minimal(base_size = 16)
```

```{r}
bootstrap_distribution <- GLP_df |>
  filter(group == "treatment") |>
  specify(response = weight_loss) |>
  generate(reps = 1337, type = "bootstrap") |>
  calculate(stat = "mean")

bootstrap_distribution |>
  visualize() +
  shade_ci(endpoints = bootstrap_distribution |>
             get_ci(level = 0.95, type = "percentile"),
           color = princeton_black,
           fill = princeton_orange) +
  labs(title = "GLP Clinical Trials",
       subtitle = "treatment group",
       caption = "(hypothetical data)",
       x = "weight loss", y = "number of patients") +
  theme_minimal(base_size = 16)
```


## Hypothesis Testing

### Null Distribution

Use `infer` package code to perform NHST to test the claim (among those in the treatment group)

* null hypothesis: male patients have at least as much weight loss than female patients
* alternative hypothesis: male patients have less weight loss than female patients

```{r}
# build a null distribution
null_distribution <- GLP_df |>
  filter(group == "treatment") |>
  specify(formula = weight_loss ~ sex) |> 
  hypothesize(null = "independence") |> 
  generate(reps = 1000, type = "permute") |> 
  calculate(stat = "diff in means", order = c("male", "female"))
```

```{r}
# compute the observed difference in means
obs_diff_means <- GLP_df |>
  filter(group == "treatment") |>
  specify(formula = weight_loss ~ sex) |> 
  calculate(stat = "diff in means", order = c("male", "female"))
  
# visualize the null distribution and shade in the p-value
null_distribution |>
visualize(bins = 10) + 
  shade_p_value(obs_stat = obs_diff_means, direction = "less") +
  theme_minimal()
```

```{r}
null_distribution |>
  get_p_value(obs_stat = obs_diff_means, direction = "less")
```

### Boxplots

Create a side-by-side boxplot to conduct a two-sided NHST for weight loss between the control and treatment groups.

```{r}
GLP_df |>
  ggplot(aes(x = group, y = weight_loss,
             fill = group)) +
  geom_boxplot() +
  geom_signif(
    comparisons = list(c("control", "treatment")),
    map_signif_level = TRUE
  ) +
  labs(title = "GLP Clinical Trials",
       subtitle = "(toward publisable results)",
       caption = "(hypothetical data)",
       x = "", y = "weight loss (pounds)") +
  theme_minimal()
```


## Machine Learning

```{r}
GLP_ml <- GLP_df |>
  select(-c(weight_before, weight_after))

data_split <- initial_split(GLP_ml)
data_train <- training(data_split)
data_fold <- vfold_cv(data_train)
lambda_grid <- grid_regular(penalty(range = c(-3, 3)),
                            levels = 50)
```

### Ridge Regression

```{r}
#| message: false
#| warning: false

ml_pipeline_fit <- linear_reg(mixture = 0, penalty = 0) |> 
  set_engine("glmnet") |>
  fit(weight_loss ~ ., data = GLP_ml)
ml_pipeline_recipe <- recipe(weight_loss ~ ., data = GLP_ml) |>
  step_novel(all_nominal_predictors()) |>
  step_dummy(all_nominal_predictors()) |> 
  step_zv(all_predictors()) |>
  step_normalize(all_numeric_predictors())
ml_pipeline_spec <- linear_reg(mixture = 0, penalty = tune()) |>
  set_engine("glmnet")
ml_pipeline_workflow <- workflow() |> 
  add_recipe(ml_pipeline_recipe) |>
  add_model(ml_pipeline_spec)

ml_pipeline_grid <- tune_grid(
  ml_pipeline_workflow,
  resamples = data_fold, 
  grid = lambda_grid
)
```

```{r}
best_penalty <- select_best(ml_pipeline_grid, 
                            metric = "rmse") |>
  pull(penalty)
predictions <- predict(ml_pipeline_fit, 
                       new_data = GLP_df |>
                         select(-weight_loss),
                       penalty = best_penalty)
true_values <- GLP_df  |> select(weight_loss)
n <- GLP_df  |> drop_na(weight_loss) |> nrow()
RMSE <- sqrt((1/n)*sum((true_values - predictions)^2))
print(RMSE)
```

### Lasso Regression

```{r}
#| message: false
#| warning: false

ml_pipeline_fit <- linear_reg(mixture = 1, penalty = 0) |> 
  set_engine("glmnet") |>
  fit(weight_loss ~ ., data = GLP_ml)
ml_pipeline_recipe <- recipe(weight_loss ~ ., data = GLP_ml) |>
  step_novel(all_nominal_predictors()) |>
  step_dummy(all_nominal_predictors()) |> 
  step_zv(all_predictors()) |>
  step_normalize(all_numeric_predictors())
ml_pipeline_spec <- linear_reg(mixture = 1, penalty = tune()) |>
  set_engine("glmnet")
ml_pipeline_workflow <- workflow() |> 
  add_recipe(ml_pipeline_recipe) |>
  add_model(ml_pipeline_spec)

ml_pipeline_grid <- tune_grid(
  ml_pipeline_workflow,
  resamples = data_fold, 
  grid = lambda_grid
)
```

```{r}
best_penalty <- select_best(ml_pipeline_grid, 
                            metric = "rmse") |>
  pull(penalty)
predictions <- predict(ml_pipeline_fit, 
                       new_data = GLP_df |>
                         select(-weight_loss),
                       penalty = best_penalty)
true_values <- GLP_df  |> select(weight_loss)
n <- GLP_df  |> drop_na(weight_loss) |> nrow()
RMSE <- sqrt((1/n)*sum((true_values - predictions)^2))
print(RMSE)
```


### Clustering

```{r}
# cross-validation folds
ml_pipeline_cv <- vfold_cv(GLP_ml, v = 5)

# specification
ml_pipeline_spec <- k_means(num_clusters = tune())

# recipe
ml_pipeline_recipe <- recipe(~weight_loss, data = GLP_ml) |>
  step_zv() |>
  step_normalize()

# workflow
ml_pipeline_workflow <- workflow(ml_pipeline_recipe, ml_pipeline_spec)

# parameter grid
kvals_grid <- grid_regular(num_clusters(), levels = 10)

# tuning
tune_results <- tune_cluster(
  ml_pipeline_workflow,
  resamples = ml_pipeline_cv,
  grid = kvals_grid,
  metrics = cluster_metric_set(sse_within_total, sse_total, sse_ratio)
)
```

```{r}
tune_results |>
  collect_metrics() |>
  filter(.metric == "sse_ratio") |>
  ggplot(aes(x = num_clusters, y = mean)) +
  geom_point(aes(x = factor(num_clusters), y = mean),
             size = 3) +
  geom_line() +
  labs(title = "Scree Plot",
       subtitle = "Aiming to choose the number of clusters",
       caption = "SML 201",
       x = "number of clusters",
       y = "within/total SSE ratio") +
  theme_minimal(base_size = 14)
```


### Visualization

Create a side-by-side arrangements of scatterplots

* horizontal: `weight_before`
* vertical: `weight_after`

1. color coded by `group`
2. color coded by cluster assignment

```{r}
cluster_fit <- k_means(num_clusters = 2) |>
  set_engine("stats") |>
  fit(~ weight_loss, data = GLP_df)
GLP_clust <- GLP_df |>
  mutate(cluster_num = extract_cluster_assignment(cluster_fit) |>
           pull())
```

```{r}
p1 <- GLP_clust |>
  ggplot(aes(x = weight_before, y = weight_after)) +
  geom_point(aes(color = group),
             alpha = 0.5, size = 3) +
  labs(title = "GLP Clinical Trials",
       subtitle = "experiment groups",
       x = "weight before (pounds)",
       y = "weight after (pounds)") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r}
p2 <- GLP_clust |>
  ggplot(aes(x = weight_before, y = weight_after)) +
  geom_point(aes(color = cluster_num),
             alpha = 0.5, size = 3) +
  labs(title = "GLP Clinical Trials",
       subtitle = "k = 2 clusters",
       x = "weight before (pounds)",
       y = "weight after (pounds)") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r}
#patchwork
p1 + p2
```

::: {.callout-warning}
## DCP3
:::


# Coda

::: {.callout-note}

## Course Learning Outcomes

By the end of this course, students will be able to:

1. Implement data science and statistical concepts and methods to data sets and real-world scenarios
2. Utilize software to load files, perform data wrangling, and produce visualizations
3. Compute results from data analyses and express the findings to academic and scientific communication audiences
4. Discuss ethical and societal impact of data science projects and equitable practices

:::



# Quo Vadimus?

:::: {.columns}

::: {.column width="40%"}
* CLO Assessment (survey)
* Exam 2 (April 23)
  * check location and other info in Canvas
* Project 3
  * assigned: April 27
  * due: May 11
:::

::: {.column width="10%"}
	
:::

::: {.column width="50%"}
{{< video https://www.youtube.com/watch?v=LXD4MGINIzQ >}}

* last-minute exam advice from 5 Second Films ([YouTube](https://www.youtube.com/watch?v=LXD4MGINIzQ))
:::

::::


# Footnotes

::: {.callout-note collapse="true"}
## (optional) Additional Resources



:::

::: {.callout-note collapse="true"}
## Session Info

```{r}
sessionInfo()
```
:::





:::: {.columns}

::: {.column width="45%"}
	
:::

::: {.column width="10%"}
	
:::

::: {.column width="45%"}

:::

::::

::::: {.panel-tabset}



:::::